An Analysis of the Use of Centipawn Loss to Detect Cheating in Chess

### 2022-11-13 Manuscript/manuscript.tex
added centipawn explanation to intro

% Go into detail about why this was such a shock. Include backgrounds of Carlsen and Niemann.

Carlsen didn't immediately state the reason as to why he withdrew at first, though many correctly assumed that he was holding his tongue in making any sort of public accusation to avoid a potential retaliatory lawsuit for slander.

It should be noted that Carlsen played Niemann again two weeks later in the Julius Baer Generation Cup, but resigned after his second move, perpetuating rumors of the scandal even further. 
This has become a matter of significant speculation both inside and outside the chess world given the nature of the situation and the people involved.

@@

The specific allegations related to how Niemann cheated OTB are unknown. Also, it would be very hard to prove cheating by simply looking at moves that are made in a single game. But what if you had access to hundreds or even thousand of games from a certain individual over a long stretch of their career?
Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different. Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding “easy” vs. “difficult” moves.

>>>>

Carlsen didn't immediately state the reason as to why he withdrew at first, though many correctly assumed that he was holding his tongue in making any sort of public accusation to avoid a potential retaliatory lawsuit for slander. It should be noted that Carlsen played Niemann again two weeks later in the Julius Baer Generation Cup, but resigned after his second move, perpetuating rumors of the scandal even further. This has become a matter of significant speculation both inside and outside the chess world given the nature of the situation and the people involved.

@@

The specific allegations related to how Niemann cheated OTB are unknown. Also, it would be very hard to prove cheating by simply looking at moves that are made in a single game. But what if you had access to hundreds or even thousand of games from a certain individual over a long stretch of their career?
Although cheating in online chess should be, in principle, at least as hard to detect as with cheating OTB, with the added disadvantage that you cannot enforce the disallowance of electronic devices, in reality it is quite different. Chess.com, which is used by millions of players has released an expose on their cheat detection methods, specifically with regard to Hans Niemann, who they allege has cheating in many games. The methods used include (1) comparing moves made to engine-recommended moves, (2) comparing player past performances to their historical strength, (3) comparing a player's performance to comparable peers, (4) looking at behavioral factors, and (5) reviewing time usage for finding "easy" vs. "difficult" moves.

Item 2 requires an assessment of move quality, which is usually done in centipawns. The centipawn is a unit of measure used in chess to quantify advantage. A centipawn is equal to 1/100 of a pawn and, although these values play no formal role in the game, they are useful to players and chess computers alike in order to evaluate positions. Centipawn loss is a calculation and numerical score given by a chess engine as the difference between the move an individual actually plays and the "strongest" move available at the time as computed by the engine. Therefore, average centipawn loss would be a calculation of the average loss incurred by each move over any particular game.
